On the power of conditional independence testing under model-X \support{EK was partially supported by NSF DMS-2113072.}

### 2022-11-21 arxiv-v5/response_to_reviews.tex

We thank the referees for their constructive comments, which we have incorporated into a further improved manuscript. The revisions includes the following updates:

\paragraph{Updated Theorem 1.}

Proposition 1 in the previous version of the manuscript stated that any valid MX CI test must also be conditionally valid. Unfortunately, we discovered a mistake in our proof of this fact, and subsequently found counterexamples. Therefore, we have removed Proposition 1 from the paper and formulated Theorem 1 as a statement about the most powerful conditionally valid test (and in particular the most powerful CRT) rather than the most powerful MX CI test overall. We leave it as an open question whether the most powerful MX CI test is in fact a CRT.

\paragraph{Added references to recent papers inspired by our results.} Recently, Gr\"unwald et al (2022) proposed an extension of the CRT to the sequential setting. They proposed a test relying on a likelihood-based statistic, citing inspiration from our Theorem 1. Similarly, Spector and Fithian (2022) proposed a more powerful version of knockoffs based on a \textit{masked likelihood ratio} statistic, citing inspiration from our Theorem 5. We have added references to these two papers to our revised manuscript to demonstrate how our work has inspired new model-X methodologies.

@@

	We have dropped the $\alpha$ from $C_\alpha$, noting that the dependence on $\alpha$ is implicit.

	\item \textsl{I would suggest changing the $g'_n$ notation as it may be mistaken for a derivative.}

	We have replaced $g'_n$ by $\bar g_n$ to avoid this confusion.

	\item \textsl{The moment conditions in Thm 4 seem unnecessarily strong. I don't think it is critical to weaken them, but for instance I think one does not need a second moment condition for a triangular array weak law of large numbers to hold. 1+$\delta$ absolute moments should suffice (e.g. lemma 19 of https://arxiv.org/pdf/1804.07203.pdf). Similarly the application of the Lyapunov CLT can use an arbitrary $\delta > 0$ rather than $\delta=1$. Perhaps these sorts of observations can slightly weaken the moment conditions.}

	We agree with this observation, and we have added a footnote to page 12 stating that
	\begin{quote}
	\textcolor{blue}{``The exponents in these moment conditions can be relaxed from 4 to $2+\delta$, in particular, requiring an appropriate triangular array weak law of large numbers with 1+$\delta$ moments. This slight weakening of moment conditions requires significantly more technical effort, so is omitted for simplicity since it does not alter the main takeaway messages of our analysis.''}
	\end{quote}
	At this stage, we have not updated the actual statements and proofs of our theoretical results to reflect this change, in part because we could not find a reference (either textbook or paper) stating the triangular array weak law of large numbers with $1+\delta$ moments. In particular, note that the result of Shah and Peters cited by the referee is not in the triangular array setup. Therefore, we feel the slight relaxation of our moment assumptions may not be worth the extra effort required. % If the referees disagree, we would be willing to add a statement and proof of the aforementioned weak law of large numbers and update our theorem statements and proofs accordingly.

@@

\noindent We agree that Section 3 still seems like a detour from our primary focus of studying the power of the CRT. To this end, we have moved the former Section 3.4 (the numerical simulation for the MX(2) $F$-test) to Appendix B. The MX(2) $F$-test and the question of robustness are not our primary focus, so we hope relegating the simulation to the appendix makes the main text more focused and streamlined. Furthermore, we have updated the last paragraph of Section 3 for a smoother transition to Section 4:

\begin{quote}
	\textcolor{blue}{``Having found that the dCRT is a natural test to apply for power against semiparametric alternatives, and that this test is equivalent to the simpler MX(2) $F$-test, we are ready to study the relationship between the power of the dCRT and the quality of the underlying machine learning procedure.''}
\end{quote}

\textsl{Regarding theme (1), Section 2.1 is interesting but very unsurprising. I would almost say Proposition  1 was not stated explicitly in prior literature because it is obvious. The authors seem to admit so in the last paragraph of Section 2.1, but perhaps this could have been pointed out sooner. My only technical concern here is that there may be a typo, or perhaps just some ambiguity, in the notation of its proof. I was expecting to see a delta function in the integrand. It could be a typo, or it could be that I misunderstood something in the notation of the proof (I’m quite sure the result is correct). In any case, Section 1 is valuable as a prelude to Section 2.2, which contains the less obvious Theorem 1.}
\\

\noindent We appreciate the referee's suggestion for us to check the proof of Proposition 1. As it turns out, the proof was in fact wrong (though for a different reason than the referee suggested). Furthermore, we found the proposition itself to be false; please see the overview of the revisions (Section~\ref{sec:overview-of-revisions}) for more on this. We have rephrased that part of the text accordingly. Despite our earlier error and current correction, we at least hope that the reviewer may acknowledge that simple claims are not necessarily obvious ones. Nevertheless, we view Section 2.1 as primarily setting the stage for the optimality result in Section 2.2, rather than a presentation of new results.
\\

@@

\textsl{Theorem 1 is also very intuitive, as it follows quite directly from the Neyman-Pearson lemma, but it is definitely worth stating explicitly. My main concern with theme (1) is that this line of inquiry seems to terminate abruptly and prematurely after the statement of Corollary 1. What should we do with this Neyman-Pearson result? Does it have any practical relevance (as we don’t deal with point alternatives in practice)? Does it really answer any questions about which statistics should be used in practice? Does it provide “quantitative explanations for empirically observed phenomena and novel insights to guide the design of MX methodology”, as we were promised in the abstract?} 
\\

\noindent Our primary theme of power of MX CI tests is addressed not only in Section 2 (finite-sample power of CRT), but also in Section 4 (asymptotic power of CRT) and Section 5 (finite-sample power of knockoffs). As far as the consequences of the Neyman-Pearson result in Section 2, we discuss in the paragraph following Corollary 1 that this result provides evidence in favor of likelihood- or loss-based test statistics, and discuss other advantages of using such statistics. At least two recent papers have taken methodological inspiration from our optimality results (Grunwald et al, 2022 for CRT and Spector and Fithian, 2022 for knockoffs). We have acknowledged these works in this revision in Section 2:
\begin{quote}
	\textcolor{blue}{``Theorem 1 has inspired an extension of the CRT to the sequential setting using a likelihood-based test statistic, accompanied by a similar optimality result (Grunwald et al, 2022).''}
\end{quote}
and Section 5:
\begin{quote}
	\textcolor{blue}{``Recently, Theorem 5 inspired a more powerful variant of knockoffs based on \textit{masked likelihood ratio} statistics, which comes with a different kind of optimality guarantee (Spector and Fithian, 2022).''}
\end{quote}
Finally, we have updated the sentence in the abstract referenced by the referee to make it more concrete:
\begin{quote}
	``In this paper, we study the power of MX CI tests, \textcolor{blue}{yielding quantitative insights into the role of machine learning and providing evidence in favor of using likelihood-based statistics in practice.''}
\end{quote}
Here, ``quantitative insights into the role of machine learning'' refers to Theorem 4 (connecting the performance of the machine learning method to the power of the CRT) and ``evidence in favor of using likelihood-based statistics in practice'' refers to Theorem 1 (showing that the likelihood test statistic is finite-sample optimal).
\\

### 2022-11-21 publication/crtpower-ejs.tex
Updated EJS version

\documentclass[12pt]{article}

 \usepackage[vmargin=1.18in,hmargin=1.1in]{geometry}
 \usepackage{amssymb,amsfonts,graphicx,amsthm,nicefrac,mathtools,bm, bbm}
 \usepackage[hidelinks]{hyperref}
 \usepackage[normalem]{ulem} %to strike the words
 \usepackage{mathrsfs}
 \usepackage{pdflscape}

% \usepackage[numbers]{natbib}
 % \usepackage[english]{babel}
% \usepackage{times}
%  \usepackage[OT1]{fontenc}  
% \usepackage{bbm,mdframed}
 \usepackage[dvipsnames]{xcolor}
% \usepackage{xspace}
 % \usepackage{rotating,multirow}
 \usepackage{graphics,tikz}

\usepackage[]{caption}

\usepackage[linesnumbered,ruled]{algorithm2e}
\let\oldnl\nl% Store \nl in \oldnl
\newcommand{\nonl}{\renewcommand{\nl}{\let\nl\oldnl}}% Remove line number for one line
\makeatother

\usepackage[uniquename=false, uniquelist = false, url = false, doi = false, isbn = false, sorting = none, natbib = true, backend = bibtex, maxbibnames=10]{biblatex}

\addbibresource{~/.references/crtpower}
% \addbibresource{crtpower}

% \usepackage{grffile}
 % \usepackage{subcaption}
 % \usepackage{epstopdf} % for postscript graphics files

% \arxiv{arXiv:1803.06790}

 % \startlocaldefs

 %\setlength{\parindent}{10pt}
 %\setlength{\parskip}{5pt}

 %----------------------------------------------------------
 % Theorem etc macros
 %----------------------------------------------------------
 \newtheorem{theorem}{Theorem}
 \newtheorem{proposition}{Proposition}
 \newtheorem{claim}{Claim}
 \newtheorem{lemma}{Lemma}
 \newtheorem{corollary}{Corollary}
 \newtheorem{question}{Question}
 \newtheorem{assumption}{Assumption}
 \theoremstyle{definition}
 \newtheorem{definition}{Definition}
   \newtheorem{setting}{Setting}
 \theoremstyle{remark}
 \newtheorem{remark}{Remark}
 \newtheorem{example}{Example}
 % \makeatletter

 %----------------------------------------------------------
 % Reference macros
 %----------------------------------------------------------
 \newcommand{\rmkref}[1]{Remark~\ref{rmk:#1}}
 \newcommand{\assumpref}[1]{Assumption~\ref{assump:#1}}
 \newcommand{\assumpsref}[1]{Assumptions~\ref{assump:#1}}
 \newcommand{\assumpssref}[1]{\ref{assump:#1}}
 \newcommand{\figref}[1]{Figure~\ref{fig:#1}}
 \newcommand{\figsref}[1]{Figures~\ref{fig:#1}}
 \newcommand{\figssref}[1]{\ref{fig:#1}}
 \newcommand{\chapref}[1]{Chapter~\ref{chap:#1}}
 \newcommand{\secref}[1]{Section~\ref{sec:#1}}
 \newcommand{\appref}[1]{Appendix~\ref{app:#1}}
 \newcommand{\eqssref}[1]{\ref{eqn:#1}}
 \newcommand{\defref}[1]{Definition~\ref{def:#1}}
 \newcommand{\exref}[1]{Example~\ref{ex:#1}}
 \newcommand{\clmref}[1]{Claim~\ref{clm:#1}}
 \newcommand{\lemref}[1]{Lemma~\ref{lem:#1}}
 \newcommand{\lemsref}[1]{Lemmas~\ref{lem:#1}}
 \newcommand{\lemssref}[1]{\ref{lem:#1}}
 \newcommand{\propref}[1]{Proposition~\ref{prop:#1}}
 \newcommand{\propsref}[1]{Propositions~\ref{prop:#1}}
 \newcommand{\propssref}[1]{\ref{prop:#1}}
 \newcommand{\thmref}[1]{Theorem~\ref{thm:#1}}
 \newcommand{\thmwref}[1]{Theorem~#1}
 \newcommand{\thmsref}[1]{Theorems~\ref{thm:#1}}
 \newcommand{\thmssref}[1]{\ref{thm:#1}}
% \newcommand{\corref}[1]{Corollary~\ref{cor:#1}}
 \newcommand{\tabref}[1]{Table~\ref{tab:#1}}
 \newcommand{\tabsref}[1]{Tables~\ref{tab:#1}}
 \newcommand{\tabssref}[1]{\ref{tab:#1}}
 \newcommand{\pwref}[1]{p.~{#1}}
 \newcommand{\ppwref}[1]{pp.~{#1}}
 \newcommand{\eqnref}[1]{\eqref{eqn:#1}}

 \setlength{\headheight}{35pt}

 %----------------------------------------------------------
 % Other macros
 %----------------------------------------------------------
 \newcommand{\iid}[0]{i.i.d.\xspace}
 \newcommand{\dist}[2]{\mathrm{dist}\!\left({#1},{#2}\right)} % distance
 \newcommand{\one}[1]{{\mathbbm{1}}_{{#1}}}
 \newcommand{\inner}[2]{\langle{#1},{#2}\rangle} % Inner product
 \newcommand{\norm}[1]{\left\lVert{#1}\right\rVert}
 %\newcommand{\PP}[1]{\mathbb{P}\left\{{#1}\right\}} % Probability
 \newcommand{\PP}[1]{\textnormal{Pr}\!\left\{{#1}\right\}} % Probability
 %\newcommand{\Pp}[2]{\mathbb{P}_{#1}\left\{{#2}\right\}} % Probability
 \newcommand{\EE}[1]{\mathbb{E}\left[{#1}\right]} % Expectation
 \newcommand{\Ep}[2]{\mathbb{E}_{#1}\left[{#2}\right]}
 \newcommand{\EEst}[2]{\mathbb{E}\left[{#1}\ \middle| \ {#2}\right]} % Conditional expectation
 %\newcommand{\PPst}[2]{\mathbb{P}\left\{{#1}\ \middle| \ {#2}\right\}} % Conditional probability
 \newcommand{\PPst}[2]{\text{Pr}\!\left\{{#1}\ \middle| \ {#2}\right\}} % Conditional probability
 % \newcommand{\TV}[1]{\var\left({#1}\right)} % Variance
 % \newcommand{\Tp}[2]{\mathbb{V}_{#1}\left[{#2}\right]}
 % \newcommand{\TVst}[2]{\var\left({#1}\ \middle|\ {#2}\right)}
 % \renewcommand{\O}[1]{\mathcal{O}\left({#1}\right)}
 % \def\R{\mathbb{R}}
 % \def\Z{\mathbb{Z}}
 \newcommand{\ee}[1]{\mathbf{e}_{{#1}}}
 \newcommand{\ident}{\mathbf{I}}
 \newcommand{\ones}{\mathbf{1}}
 \newcommand{\zeros}{\mathbf{0}}
 \def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
 \newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
 \newcommand{\iidsim}{\stackrel{\mathrm{iid}}{\sim}}
 \newcommand{\indsim}{\stackrel{\independent}{\sim}}
 \newcommand{\notate}[1]{\textcolor{red}{\textbf{[#1]}}}
 \newcommand{\ignore}[1]{}
 \renewcommand{\Pr}[2]{\mathcal{P}_{{#1}}\left({#2}\right)} % projection
 \newcommand{\Prp}[2]{\mathcal{P}_{{#1}}^{\perp}\left({#2}\right)}
 \newcommand{\pr}[1]{\mathcal{P}_{{#1}}}
 \newcommand{\prp}[1]{\mathcal{P}_{{#1}}^{\perp}}
 \newcommand{\eps}{\epsilon}
 \newcommand{\lam}{\lambda}
 % \newcommand{\T}{\mathcal{T}}
 % \newcommand{\diff}[1]{\text{d}{#1}} % differential
 \let\emptyset\varnothing
 % \newcommand{\tf}{\widetilde{f}}
 % \newcommand{\tbeta}{\widetilde\beta}
 % \newcommand{\bhy}{\textnormal{BHY}}

@@

\title{On the power of conditional independence testing \\ under model-X}
\author{Eugene Katsevich$^1$ and Aaditya Ramdas$^{1,2}$\\
	[10pt]
	\texttt{ekatsevi@wharton.upenn.edu},
	\texttt{aramdas@stat.cmu.edu} \\
	Department of Statistics and Data Science, University of Pennsylvania$^1$\\
	Department of Statistics and Data Science, 	Carnegie Mellon University$^2$\\
	Machine Learning Department, Carnegie Mellon University$^2$
}

\maketitle
\thispagestyle{empty}

\begin{abstract}
For testing conditional independence (CI) of a response $Y$ and a predictor $X$ given covariates $Z$, the recently introduced model-X (MX) framework has been the subject of active methodological research, especially in the context of MX knockoffs and their successful application to genome-wide association studies. In this paper, we study the power of MX CI tests, yielding quantitative insights into the role of machine learning and providing evidence in favor of using likelihood-based statistics in practice. Focusing on the conditional randomization test (CRT), we find that its conditional mode of inference allows us to reformulate it as testing a point null hypothesis involving the conditional distribution of $X$. The Neyman-Pearson lemma then implies that a likelihood-based statistic yields the most powerful CRT against a point alternative. We also obtain a related optimality result for MX knockoffs. Switching to an asymptotic framework with arbitrarily growing covariate dimension, we derive an expression for the limiting power of the CRT against local semiparametric alternatives in terms of the prediction error of the machine learning algorithm on which its test statistic is based. Finally, we exhibit a resampling-free test with uniform asymptotic Type-I error control under the assumption that \textit{only the first two moments of $X$ given $Z$ are known}, a significant relaxation of the MX assumption. 
\end{abstract}

%	\begin{keyword}
%		\kwd{conditional independence testing}
%		\kwd{conditional randomization test}
%		\kwd{model-X}
%		\kwd{knockoffs}
%		\kwd{most powerful tests}
%		\kwd{local alternatives}
%		\kwd{dose response curve}
%	\end{keyword}

%\newpage 	
%\footnotesize
%\tableofcontents
%\normalsize

\color{black}

@@

%To keep an example in mind throughout the paper, consider a genome-wide association study (GWAS) based on $n$ individuals. For each individual, a disease status $\pry \in \{0,1\}$ is recorded along with the genotypes $\prx \in \{0,1,2\}$ at a single nucleotide polymorphism (SNP) of interest and $\prz \in \{0,1,2\}^p$ at all other SNPs. SNPs usually have only two possible alleles in the population (the \textit{minor allele} defined as the less common one), so the genotype at a SNP is defined as the number of minor alleles across maternal and paternal chromosomes. For example, $\prx = 1$ for an individual whose maternal and paternal chromosomes have a minor and a major allele, respectively, at the SNP of interest.

%
%consider a neuroscience experiment where we wish to determine if a particular \textit{voxel} (volume pixel) in the visual cortex responds preferentially to faces using an fMRI (functional magnetic resonance imaging) experiment involving showing a subject a sequence of $n$ images. Here, $\pry \in \mathbb R$ would be the fMRI brain recording 6 seconds after showing the image, $\prx \in \mathbb R$ would be a count of the number of faces in the image, and $\prz \in \mathbb R^{100}$ would include other visual features (counts of non-face objects, edge-detectors, color/brightness).
% \textcolor{red}{To keep an example in mind throughout the paper, consider $\pry \in \mathbb R^1$ cholesterol level, $\prx \in \{0,1,2\}^{10}$ the genotypes of an individual at 10 nearby polymorphic sites, and $\prz \in \{0,1,2\}^{500,000}$ the genotypes of the individual at other polymorphic sites across the genome.}

As formalized by Shah and Peters \cite{Shah2018}, the problem \eqref{conditional-independence} is fundamentally impossible without assumptions on the distribution $\mathcal L(\prx,\pry, \prz)$, in which case no asymptotically uniformly valid test of this hypothesis can have nontrivial power against \emph{any} alternative. In special cases, the problem is more tractable, for example if $\prz$ has discrete support, or if we were willing to make (semi)parametric assumptions on the form of $\mathcal L(\pry|\prx, \prz)$ (henceforth ``model-$Y|X$''). We will not be making such assumptions in this work.
% A complementary recent line of work has focused on shifting assumptions to the distribution $f_{\prx, \prz}$. In particular, 
Instead, we follow the lead of Candes et al. \cite{CetL16}, who proposed to avoid assumptions on $\mathcal L(\pry|\prx, \prz)$, but assume that we have access to $\mathcal L(\prx|\prz)$:\footnote{Candes et al. actually require that the full joint distribution $\mathcal L(\prx,\prz)$ is known, but this is because they also test for conditional associations between $\prz$ and $\pry$. We focus only on the relationship between $\prx$ and $\pry$ given $\prz$ and therefore require a weaker assumption.} 
\begin{equation} \label{eq:modelX}
\textit{model-X (MX) assumption}: \mathcal L(\prx| \prz)=  f^*_{\prx|\prz} \text{ for some known } f^*_{\prx|\prz}.\footnote{We implicitly assume that $\mathcal L$ has a density with respect to some dominating measure on $\mathbb R^{1+1+p}$, and that all conditional densities are well-defined almost surely. Here and throughout the paper, we identify probability distributions with their densities with respect to the appropriate dominating measure.}
\end{equation}
Candes et al.\ argue that while both model-$Y|X$ and MX are strong assumptions---especially when $p,d$ are large---in certain cases much more is known about $\prx | \prz$ than about $\pry|\prx, \prz$. In the aforementioned GWAS example, $\prx|\prz$ reflects the joint distribution of genotypes at SNPs across the genome, which is well described by hidden Markov models from population genetics \cite{SetC17}. On the other hand, the distribution $\pry|\prx,\prz$ represents the genetic basis of a complex trait, about which much less is known. In the context of (stratified) randomized experiments, the distribution $\mathcal L(\prx | \prz)$ is the propensity function \cite{Imai2004} (the analog of the propensity score for non-binary treatments \cite{Rosenbaum1983}) and is experimentally controlled. In general causal inference contexts, the MX assumption can be viewed as the assumption that the propensity function is known.

@@

Testing CI hypotheses in the MX framework has been the subject of active methodological research. The most popular methodology is MX knockoffs \cite{CetL16}. This method is based on the idea of constructing synthetic negative controls (knockoffs) for each predictor variable in a rigorous way that is based on the MX assumption; see Section~\ref{sec:knockoffs-overview} for a brief overview. Rapid progress has been made on the construction of knockoffs in various cases \cite{SetC17,Romano2019a,Bates2019,Huang2019} and on the application of this methodology to GWAS \cite{SetC17, SetS19}. The conditional randomization test (CRT) \cite{CetL16}, initially less popular than knockoffs due to its computational cost, is receiving renewed attention as computationally efficient variants are proposed, such as the holdout randomization test (HRT) \cite{Tansey2018}, the digital twin test \cite{Bates2020}, and the distilled CRT (dCRT) \cite{Liu2020}. The dCRT in particular is a promising methodology because it combines good power and computational speed; we focus on this variant of the CRT is Sections~\ref{sec:weakening} and~\ref{sec:asymptotic-power} of this paper. We introduce the general CRT methodology next, while deferring the introduction of the dCRT to Section~\ref{sec:weakening}.

We start with any test statistic $T(\srx, \sry, \srz)$ measuring the association between $\prx$ and $\pry$, given $\prz$. Usually, this statistic involves learning some estimate $\widehat f_{\pry|\prx,\prz}$ based on machine learning, e.g. the magnitude of the fitted coefficient for $\prx$ (when $\text{dim}(\prx) = 1$) in a cross-validated lasso~\cite{T96} of $\sry$ on $\srx$ and $\srz$ \cite{CetL16}. To calculate the distribution of $T$ under the null hypothesis~\eqref{conditional-independence}, first define a matrix  $\srxk \in \mathbb R^{n \times d}$, where the $i$th row $\srxk_i$ is a sample from $\mathcal L(\prx \mid \prz = \srz_i)$. In other words, for each sample $i$, resample $\srx_i$ based on its distribution conditional on the observed covariate values $Z_i$ in that sample. We then use these resamples to build a null distribution $T(\srxk, \sry, \srz)$, from which the upper quantile
\begin{equation}
C(\sry,\srz) \equiv Q_{1-\alpha}[T(\srxk, \sry, \srz)|\sry,\srz]
	\label{upper-quantile}
\end{equation}
may be extracted (the dependence on $\alpha$ left implicit), where the randomness is over the resampling distribution $\srxk |  \sry, \srz$. Finally, the CRT rejects if the original test statistic exceeds this quantile:

>>>>

\documentclass[ejs]{imsart}

\RequirePackage{amsthm,amsmath,amsfonts, amssymb, mathrsfs}
\RequirePackage[authoryear]{natbib}
\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
\RequirePackage{graphicx}
\RequirePackage{bm, bbm}

\RequirePackage[linesnumbered,ruled]{algorithm2e}
\let\oldnl\nl% Store \nl in \oldnl
\newcommand{\nonl}{\renewcommand{\nl}{\let\nl\oldnl}}% Remove line number for one line
\makeatother

% settings
\pubyear{2022}
\volume{0}
\issue{0}
\firstpage{1}
\lastpage{8}
\arxiv{2005.05506}

\startlocaldefs
\numberwithin{equation}{section}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
 \newtheorem{proposition}{Proposition}
 \newtheorem{claim}{Claim}
 \newtheorem{lemma}{Lemma}
 \newtheorem{corollary}{Corollary}
 \newtheorem{question}{Question}
 \newtheorem{assumption}{Assumption}
 \theoremstyle{definition}
 \newtheorem{definition}{Definition}
   \newtheorem{setting}{Setting}
 \theoremstyle{remark}
 \newtheorem{remark}{Remark}
 \newtheorem{example}{Example}

\newcommand{\iid}[0]{i.i.d.\xspace}
 \newcommand{\dist}[2]{\mathrm{dist}\!\left({#1},{#2}\right)} % distance
 \newcommand{\one}[1]{{\mathbbm{1}}_{{#1}}}
 \newcommand{\inner}[2]{\langle{#1},{#2}\rangle} % Inner product
 \newcommand{\norm}[1]{\left\lVert{#1}\right\rVert}
 \newcommand{\PP}[1]{\textnormal{Pr}\!\left\{{#1}\right\}} % Probability
 \newcommand{\EE}[1]{\mathbb{E}\left[{#1}\right]} % Expectation
 \newcommand{\Ep}[2]{\mathbb{E}_{#1}\left[{#2}\right]}
 \newcommand{\EEst}[2]{\mathbb{E}\left[{#1}\ \middle| \ {#2}\right]} % Conditional expectation
 \newcommand{\PPst}[2]{\text{Pr}\!\left\{{#1}\ \middle| \ {#2}\right\}} % Conditional probability
 \newcommand{\ee}[1]{\mathbf{e}_{{#1}}}
 \newcommand{\ident}{\mathbf{I}}
 \newcommand{\ones}{\mathbf{1}}
 \newcommand{\zeros}{\mathbf{0}}
 \def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
 \newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
 \newcommand{\iidsim}{\stackrel{\mathrm{iid}}{\sim}}
 \newcommand{\indsim}{\stackrel{\independent}{\sim}}
 \newcommand{\notate}[1]{\textcolor{red}{\textbf{[#1]}}}
 \newcommand{\ignore}[1]{}
 \renewcommand{\Pr}[2]{\mathcal{P}_{{#1}}\left({#2}\right)} % projection
 \newcommand{\Prp}[2]{\mathcal{P}_{{#1}}^{\perp}\left({#2}\right)}
 \newcommand{\pr}[1]{\mathcal{P}_{{#1}}}
 \newcommand{\prp}[1]{\mathcal{P}_{{#1}}^{\perp}}
 \newcommand{\eps}{\epsilon}
 \newcommand{\lam}{\lambda}

 \let\emptyset\varnothing

@@

\begin{frontmatter}
\title{On the power of conditional independence testing under model-X \support{EK was partially supported by NSF DMS-2113072.}}
\runtitle{Power of model-X testing}

\begin{aug}
\author{\fnms{Eugene} \snm{Katsevich}\ead[label=e1]{ekatsevi@wharton.upenn.edu}}
\and
\author{\fnms{Aaditya} \snm{Ramdas}\ead[label=e2]{aramdas@stat.cmu.edu}}

\address{Department of Statistics and Data Science, University of Pennsylvania\\
Department of Statistics and Data Science, Carnegie Mellon University\\
Machine Learning Department, Carnegie Mellon University \\
\printead{e1,e2}}
\runauthor{Katsevich and Ramdas}

\end{aug}

\begin{abstract}
       For testing conditional independence (CI) of a response $Y$ and a predictor $X$ given covariates $Z$, the model-X (MX) framework has been the subject of active methodological research, especially in the context of MX knockoffs and their application to genome-wide association studies. In this paper, we study the power of MX CI tests, yielding quantitative insights into the role of machine learning and providing evidence in favor of using likelihood-based statistics in practice. Focusing on the conditional randomization test (CRT), we find that its conditional mode of inference allows us to reformulate it as testing a point null hypothesis involving the conditional distribution of $X$. The Neyman-Pearson lemma implies that a likelihood-based statistic yields the most powerful CRT against a point alternative. We obtain a related optimality result for MX knockoffs. Switching to an asymptotic framework with arbitrarily growing covariate dimension, we derive an expression for the power of the CRT against local semiparametric alternatives in terms of the prediction error of the machine learning algorithm on which its test statistic is based. Finally, we exhibit a resampling-free test with uniform asymptotic Type-I error control under the assumption that only the first two moments of $X$ given $Z$ are known.
\end{abstract}

\begin{keyword}[class=MSC]
\kwd[Primary ]{60K35}
\kwd{60K35}
\kwd[; secondary ]{60K35}
\end{keyword}

\begin{keyword}
\kwd{Model-X}
\kwd{Conditional independence testing}
\kwd{Conditional inference}
\kwd{Neyman-Pearson}
\kwd{Power analysis}
\end{keyword}
\tableofcontents
\end{frontmatter}

@@

As formalized by \citet{Shah2018}, the problem \eqref{conditional-independence} is fundamentally impossible without assumptions on the distribution $\mathcal L(\prx,\pry, \prz)$, in which case no asymptotically uniformly valid test of this hypothesis can have nontrivial power against \emph{any} alternative. In special cases, the problem is more tractable, for example if $\prz$ has discrete support, or if we were willing to make (semi)parametric assumptions on the form of $\mathcal L(\pry|\prx, \prz)$ (henceforth ``model-$Y|X$''). We will not be making such assumptions in this work. Instead, we follow the lead of \citet{CetL16}, who proposed to avoid assumptions on $\mathcal L(\pry|\prx, \prz)$, but assume that we have access to $\mathcal L(\prx|\prz)$:\footnote{Candes et al. actually require that the full joint distribution $\mathcal L(\prx,\prz)$ is known, but this is because they also test for conditional associations between $\prz$ and $\pry$. We focus only on the relationship between $\prx$ and $\pry$ given $\prz$ and therefore require a weaker assumption.} 
\begin{equation} \label{eq:modelX}
\textit{model-X (MX) assumption}: \mathcal L(\prx| \prz)=  f^*_{\prx|\prz} \text{ for some known } f^*_{\prx|\prz}.\footnote{We implicitly assume that $\mathcal L$ has a density with respect to some dominating measure on $\mathbb R^{1+1+p}$, and that all conditional densities are well-defined almost surely. Here and throughout the paper, we identify probability distributions with their densities with respect to the appropriate dominating measure.}
\end{equation}
Candes et al.\ argue that while both model-$Y|X$ and MX are strong assumptions---especially when $p,d$ are large---in certain cases much more is known about $\prx | \prz$ than about $\pry|\prx, \prz$. In the aforementioned GWAS example, $\prx|\prz$ reflects the joint distribution of genotypes at SNPs across the genome, which is well described by hidden Markov models from population genetics \citep{SetC17}. On the other hand, the distribution $\pry|\prx,\prz$ represents the genetic basis of a complex trait, about which much less is known. In the context of (stratified) randomized experiments, the distribution $\mathcal L(\prx | \prz)$ is the propensity function \citep{Imai2004} (the analog of the propensity score for non-binary treatments, \cite{Rosenbaum1983}) and is experimentally controlled. In general causal inference contexts, the MX assumption can be viewed as the assumption that the propensity function is known.

@@

Testing CI hypotheses in the MX framework has been the subject of active methodological research. The most popular methodology is MX knockoffs \citep{CetL16}. This method is based on the idea of constructing synthetic negative controls (knockoffs) for each predictor variable in a rigorous way that is based on the MX assumption; see Section~\ref{sec:knockoffs-overview} for a brief overview. Rapid progress has been made on the construction of knockoffs in various cases \citep{SetC17,Romano2019a,Bates2019,Huang2019} and on the application of this methodology to GWAS \citep{SetC17, SetS19}. The conditional randomization test (CRT) \citep{CetL16}, initially less popular than knockoffs due to its computational cost, is receiving renewed attention as computationally efficient variants are proposed, such as the holdout randomization test (HRT) \citep{Tansey2018}, the digital twin test \citep{Bates2020}, and the distilled CRT (dCRT) \citep{Liu2020}. The dCRT in particular is a promising methodology because it combines good power and computational speed; we focus on this variant of the CRT is Sections~\ref{sec:weakening} and~\ref{sec:asymptotic-power} of this paper. We introduce the general CRT methodology next, while deferring the introduction of the dCRT to Section~\ref{sec:weakening}.

We start with any test statistic $T(\srx, \sry, \srz)$ measuring the association between $\prx$ and $\pry$, given $\prz$. Usually, this statistic involves learning some estimate $\widehat f_{\pry|\prx,\prz}$ based on machine learning, e.g. the magnitude of the fitted coefficient for $\prx$ (when $\text{dim}(\prx) = 1$) in a cross-validated lasso~\citep{T96} of $\sry$ on $\srx$ and $\srz$ \citep{CetL16}. To calculate the distribution of $T$ under the null hypothesis~\eqref{conditional-independence}, first define a matrix  $\srxk \in \mathbb R^{n \times d}$, where the $i$th row $\srxk_i$ is a sample from $\mathcal L(\prx \mid \prz = \srz_i)$. In other words, for each sample $i$, resample $\srx_i$ based on its distribution conditional on the observed covariate values $Z_i$ in that sample. We then use these resamples to build a null distribution $T(\srxk, \sry, \srz)$, from which the upper quantile
\begin{equation}
C(\sry,\srz) \equiv Q_{1-\alpha}[T(\srxk, \sry, \srz)|\sry,\srz]
	\label{upper-quantile}
\end{equation}
may be extracted (the dependence on $\alpha$ left implicit), where the randomness is over the resampling distribution $\srxk |  \sry, \srz$. Finally, the CRT rejects if the original test statistic exceeds this quantile:

\begin{equation}
	\label{eq:randomized-CRT}
	\phi^{\CRT}_T(\srx,\sry,\srz) \equiv 
	\begin{cases}
		1, \quad &\text{if }  T(\srx, \sry, \srz) > C(\sry,\srz); \\
		\gamma, \quad &\text{if }  T(\srx, \sry, \srz) = C(\sry,\srz); \\
		0, \quad &\text{if }T(\srx, \sry, \srz) < C(\sry,\srz).
	\end{cases}
\end{equation}
In order to accommodate discreteness, the CRT makes a randomized decision $\gamma$ when $T(\srx, \sry, \srz) = C(\sry,\srz)$ so that the size of the test is exactly $\alpha$. In practice, the threshold $C(\sry,\srz)$ is approximated by computing $T(\srxk^b, \sry, \srz)$ for a large number $B$ of Monte Carlo resamples $\srxk^b \sim \srx|\srz$. For the sake of clarity, this paper considers only the ``infinite-$B$" version of the CRT as defined by~\eqref{upper-quantile} and \eqref{eq:randomized-CRT}. In the causal inference setting, the CRT can be viewed as a variant of Fisher's exact test for randomized experiments that incorporates strata of covariates \citep{Zheng2008,Hennessy2016}, basing inference on rerandomizing the treatment to the units.

The CI testing problem under MX has benefited from several methodological innovations, but fundamental questions regarding power and optimality have received less attention. Therefore, in this paper we address the following two primary questions:
\begin{enumerate}
	\item[Q1.] Are there ``optimal" test statistics for MX methods, in any sense?
	\item[Q2.] What is the precise connection between the performance of the machine learning algorithm and the power of the resulting MX method?
\end{enumerate}
To the best of our knowledge, Q1 has not been considered before, while Q2 has only been indirectly addressed in the context of lasso-based knockoffs \citep{Weinstein2017, Liu2019, Fan2020, Weinstein2020} and CRT \citep{Wang2020b, Celentano2020}. The present paper complements these existing works by considering arbitrary machine learning methods. We summarize our findings next.
